A new method for genome-wide marker development and typing holds great promise for primate population genetics

### 2012-12-14 paper/paper.tex
Done except for Andy's edits

In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research. During the first decades of the discipline, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest. The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011). Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or a few microsatellites (Ting and Sterner, 2012). Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006). Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers (Edwards, 2009).

@@

	We digested genomic DNA from 5 primates with \emph{PspXI} (New England Biolabs) and used it to create a multiplexed RAD tag library. Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3'). Individual-specific barcodes contained in the P1 adapter differed by at least \textcolor{red}{three} nucleotides. We chose \emph{PspXI} based on the results of \emph{in silico} digestion of the human, rhesus macaque, and baboon reference genomes using custom Perl scripts. We sequenced the prepared library as one 150-cycle paired-end run and one 150-cycle single-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites. Other individuals were sequenced alongside those of the present study. Sequences are available to download from the NCBI Short Read Archive (accession number \textcolor{red}{SRAXXXXXX.X}).

@@

Despite the great potential of the application of RAD sequencing in primatology, we also need to point out some possible limitations of this technique. Possibly, the main constraint of employing RAD-seq on a large scale within primates is related to the need of high quality samples in order to build the library. In this study we employed high quality DNA, extracted from tissue or blood. However, most molecular primatologists are limited in their use of invasive samples, and more often rely on low quality samples such as hair or feces. \textcolor{red}{[I'd like to cut this, since I don't want to hint that we're working on techniques to get RAD tags out of feces] Although not yet available, in theory, RAD-seq protocols using non invasive samples could be developed. In a recent study, Perry and colleagues (2010) presented a genomic-scale capture protocol to obtain endogenous DNA from primate fecal samples. Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011; Guschanski et al., in press) or even fossils (Burbano et al., 2010; Krause et al., 2010). It is likely that similar methods can be developed soon allowing primatologists to employ RAD-seq also on low quality DNA samples, such as feces or museum specimens.}

Another possible limitation of the RAD-seq approach is the time scale of its application. RAD-seq data in fact might be not suitable for comparing taxa very distantly related (Rubin et al., 2012). In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a major determinant of the success of the RAD method (Rubin et al., 2012). The deep divergences between taxa in fact decreases the amount of discoverable RAD loci for two main reason: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012). This correlation between accuracy and divergence time either reduces the number of orthologus loci available for phylogenetic reconstruction or increase the amount of missing data; both scenarios can affect phylogenetic performance reducing the support values in many nodes or supporting different topologies. However, despite this drawback, Rubin and colleagues were successful in reconstructing the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya. This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).

This study illustrates the value of RAD-seq approach in discovering a large number of independent SNPs that can be used to address several questions in primatology, ranging from population genomics to phylogenetics. Our preliminary study on primates show the feasibility of this techniques across the primate order, even when nearby reference genomes are not available. Further developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.

@@

\textcolor{red}{Leakey Foundation Grant} The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.

>>>>

In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research (Avise, 1994). During the first decades of the discipline, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest. The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011). Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or a few microsatellites (Ting and Sterner, 2012). Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006). Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers (Edwards, 2009).

@@

	We digested genomic DNA from 5 primates with \emph{PspXI} (New England Biolabs) and used it to create a multiplexed RAD tag library. Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter bottom oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'- /5Phos/GAT CGG AAG AGC GGT TCA GCA GGA ATG CCG AGA CCG ATC AGA ACA A-3'; P2\_bottom: 5'- CAA GCA GAA GAC GGC ATA CGA GAT CGG TCT CGG CAT TCC TGC TGA ACC GCT CTT CCG ATC*T -3'). Individual-specific barcodes contained in the P1 adapter differed by at least three nucleotides. We chose \emph{PspXI} based on the results of \emph{in silico} digestion of the human, rhesus macaque, and baboon reference genomes using custom Perl scripts. We sequenced the prepared library as one 150-cycle paired-end run and one 150-cycle single-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites. Other individuals were sequenced alongside those of the present study. Sequences are available to download from the NCBI Short Read Archive (accession number SRAXXXXXX.X).

@@

Despite the great potential of the application of RAD sequencing in primatology, we also need to point out some possible limitations of this technique. Possibly, the main constraint of employing RAD-seq on a large scale within primates is related to the need of high quality samples in order to build the library. In this study we employed high quality DNA, extracted from tissue or blood. However, most molecular primatologists are limited in their use of invasive samples, and more often rely on low quality samples such as hair or feces. Although not yet available, in theory, RAD-seq protocols using non invasive samples could be developed. In a recent study, Perry and colleagues (2010) presented a genomic-scale capture protocol to obtain endogenous DNA from primate fecal samples. Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011; Guschanski et al., in press) or even fossils (Burbano et al., 2010; Krause et al., 2010).

Another possible limitation of the RAD-seq approach is the time scale of its application. RAD-seq data in fact might be not suitable for comparing very distantly related taxa (Rubin et al., 2012). In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a critical factor in the success of the RAD method (Rubin et al., 2012). The deep divergences between taxa in fact decreases the amount of discoverable RAD loci for two main reason: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012). This correlation between accuracy and divergence time either reduces the number of orthologus loci available for phylogenetic reconstruction or increase the amount of missing data; both scenarios can affect phylogenetic performance reducing the support values in many nodes or supporting different topologies. However, despite this drawback, Rubin and colleagues were successful in reconstructing the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya. This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85 Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).

This study illustrates the value of RAD-seq approach in discovering a large number of independent SNPs that can be used to address several questions in primatology, ranging from population genomics to phylogenetics. Our preliminary study on primates show the feasibility of this techniques across the primate order, even when nearby reference genomes are not available. Future developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.

@@

The present study was supported by a Leakey Foundation General Grant. The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.
